---
title: "Data Wrangling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Import

## Importing files into a dataframe
```{r error = TRUE, warning = FALSE}
read.csv(file = "filename.csv",
         header = T,
         row.names = 1,
         na.strings = ,
         stringsAsFactors = F)
```

## Tibbles

```{r, warning = FALSE, message=FALSE}
library(dplyr)
tibble("name" = c("me", "you", "I"),
       "age" = c(22, 23, 24),
       "gender" = c("m", "f", "m"))
```
## Tibble vs. Data Frame

1. tibble() never adjusts the names of variables.

```{r}
mytibble <- tibble("first name" = c("me", "you", "I"))
names(mytibble)
mydata <- data.frame("first name" = c("me", "you", "I"))
names(mydata)
```

2. tibble() evaluates arguments lazily and sequentially.

```{r error=TRUE}
data.frame(x = 1:5, y = x^2)
tibble(x = 1:5, y = x^2)
```

3. Tibbles are quite strict about subsetting. [i,j] always
returns another tibble. Contrast this with a data frame,
where [i,j] may return a data frame or a vector.

```{r}
tb <- tibble("name" = c("me", "you", "I"),
       "age" = c(22, 23, 24),
       "gender" = c("m", "f", "m"))

df <- data.frame("name" = c("me", "you", "I"),
       "age" = c(22, 23, 24),
       "gender" = c("m", "f", "m"))

class(tb[ , 2])
class(df[ , 2])
```

2. To extract a column as a vector object, use $. 
Tibbles are also stricter with $. It doesn’t allow partial naming.

```{r}
tb <- tibble("name" = c("me", "you", "I"),
       "age" = c(22, 23, 24),
       "gender" = c("m", "f", "m"))

tb$age
class(tb$age)
tb$a

df <- data.frame("name" = c("me", "you", "I"),
       "age" = c(22, 23, 24),
       "gender" = c("m", "f", "m"))

df$age
class(df$age)
df$a

```

## Importing files into tibbles

```{r}
library(readr)
read_csv(file = "flat file 01.txt")
read_csv(file = "flat file 02.txt",
         na = "---",
         col_names = c("name", "sex", "age"))
```
## Parsing Factors

```{r}
sex = c("M", "F", "M", "F", "M", "M")
sex_f = parse_factor(sex, levels = c("M", "F"), ordered = F) 
str(sex_f)
```

## Parsing numbers with formats
```{r}
parse_double("1000")
parse_double("1,000")
parse_number("$1,000.00")
parse_number("The income is $1,000.00")
parse_number("$98.76")
```
## Parsing date/time

parse_datetime()
```{r}
parse_datetime("2018-09-28T1310")
parse_datetime("20180928T1310")
parse_datetime("09/28/2018")
```

parse_date()

```{r}
parse_date("2018/09/28")
parse_date("2018-09-28")
```

parse_time()

```{r}
parse_time("1:00 PM")
parse_time("13:00")
```

```{r}
parse_date("Sep-28-2018", format = "%d - %b - %Y")
parse_time("1100PM", format = "%I%M%p")
```
## Specifying column types in read_csv

```{r}
read_csv(file = "flat file 07.txt", na = "NDA",
         col_types = cols(
                 "Name" = col_character(),
                 "Sex" = col_factor(levels = c("M", "F"), ordered = F),
                 "Age" = col_integer(),
                 "Income" = col_number(),
                 "DateEntry" = col_date(format = "%m/%d/%Y")
         ))
```

## Reading Fixed Width Files

```{r}
# specify column widths
read_fwf(file = "flat file 06.txt", col_positions = 
                 fwf_cols("Name" = 10,
                          "Sex" = 3,
                          "Age" = 3,
                          "Income" = 5), skip = 1)

read_fwf(file = "flat file 06.txt", col_positions = 
                 fwf_cols("Name" = c(1,10),
                          "Sex" = c(11,13),
                          "Age" = c(14,16),
                          "Income" = c(17,21)), skip = 1)
```
## Writing tibbles to a csv file

```{r}
write_csv(mtcars, "mtcars.csv")
write_rds(mtcars, "mtcars.rds")
```

# Tidying Data

**Gather**
```{r}
library(tidyr)
table4a

data <- gather(table4a, "1999", "2000", key = "Year", value = "cases", factor_key = TRUE)
data
```

**Spread**

```{r}
table2
spread(table2, key = type, value = count)
```

**Separate**

```{r}
table3
separate(table3, col = rate, into = c("case", "population"), convert = TRUE)
separate(table3, col = rate, into = c("case", "population"), sep = "/", convert = TRUE)
```

**Unite**

```{r}
table5
data <- unite(table5, col = yr, century, year, sep = "")
separate(data, col = rate, into = c("cases", "population"), sep = "/", convert = TRUE)
```

**Duplicates in the data**

Use unique() and duplicated(). Also, from dyplyr package we have distinct().

## Transforming Data Using dplyr

**Filter**

```{r}
library(nycflights13)
flights
filter(flights, month == 2, day == 2)
```

**Arrange**

```{r}
arrange(flights, month, day, desc(dep_delay))

```


**Select**

```{r}
select(flights, -(year:dep_time))
select(flights, year)
select(flights, -c(year, dep_time, day))
```

**Mutate**

```{r}
# select columns i.e obtain relevant variables to the analysis
flight_sub <- select(flights, year:day, dep_delay, arr_delay)
flight_sub

# total gain with respect to expected travel time 
mema <- mutate(flight_sub, gain = arr_delay - dep_delay)
rename(mema, worth = gain)
transmute(flight_sub, gain = arr_delay - dep_delay)
```

**Summarize**

```{r}
flights_grp = group_by(flights, year, month)
flights_grp

summarize(flights_grp, "Avg. arrival delay" = mean(arr_delay, na.rm = T),
                        "Avg. departure delay" = mean(dep_delay, na.rm = T))


oct <- filter(flights, month == 10)
oct

oct <- group_by(oct, year)
oct

summarize(oct, "Avg. arrival delay" = mean(arr_delay, na.rm = T),
                        "Avg. departure delay" = mean(dep_delay, na.rm = T))


# using count()

by_mnth <- group_by(flights, month)
by_mnth 
count(by_mnth)

by_yr <- group_by(flights, year)
by_yr
count(by_yr) # we only have 1 year
```



