---
title: "Solutions to R for Data Science by HWickham"
date: "11/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading Packages
```{r message = F}
library(tidyverse)
library(nycflights13)
```

View data
```{r}
flights
names(flights)
names(flights)
```

## Page 49 Exercise

a. Had an arrival delay of two or more hours
```{r}
data <- select(flights, arr_delay, flight)
data <- filter(data, arr_delay >= 120)
data
```

b. Flew to Houston (IAH or HOU)
```{r}
data <- select(flights, flight, dest)
data <- filter(data, dest == "IAH"|dest == "HOU")
data
```

c. Were operated by United, American, or Delta
```{r}
data <- select(flights, flight, carrier)
data <- filter(data, carrier == "UA"|carrier == "DL"|carrier == "AA")
data

```

d. Departed in summer (July, August, and September)
```{r}
data <- select(flights, flight, month)
data <- filter(data, month %in% c(7, 8, 9))
data
```

e. Arrived more than two hours late, but didn’t leave late
```{r}
data <- select(flights, flight, arr_delay, dep_delay)
data <- filter(data, dep_delay <= 0 & arr_delay > 120)
data
```

f. Were delayed by at least an hour, but made up over 30
minutes in flight
```{r}
data <- select(flights, flight, dep_delay, arr_delay)
data <- filter(data, dep_delay >= 60 & dep_delay - arr_delay > 30)
data
```

g. Departed between midnight and 6 a.m. (inclusive)
```{r}
data <- select(flights, flight, dep_time)
data <- filter(data, dep_time == 2400 | dep_time <= 600)
data
```


2. Another useful dplyr filtering helper is between(). What does it
do? Can you use it to simplify the code needed to answer the
previous challenges?

```{r}
filter(flights, between(dep_time, 601, 2359))
filter(flights, between(month, 7, 9))
```

3. How many flights have a missing dep_time? What other variables
are missing? What might these rows represent?
```{r}
summary(flights)
```
8255 flights have a missing `dep_time`, 8255 have a missing `dep_delay`, 
8713 have a missing `arr_time`, 9430 have a missing `arr_delay`, 
and 9430 have a missing `air_time`. We can speculate that these 
are flights that failed to depart or arrive, since a flight t
hat departs normally but is then rerouted will probably have a normally recorded departure but no similar record for it's arrival. 
However, these could also just be lost data about perfectly normal flights.


4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing?
Why is FALSE & NA not missing? Can you figure out the general
rule? (NA * 0 is a tricky counterexample!)
```{r}
NA^0
NA | TRUE
FALSE & NA
NA * 0
```

NA^0 is 1 since although we dont know the number, any number raise to 0 is 1.

The logical operator | w/c stands for or returns a value true when at least 
one of the statement is true.

The logical operator & returns a valueee true iff all the statements are
true.

NA might take the value infinity so that leaves us NA result.

## Page 51 Exercise

1. How could you use arrange() to sort all missing values to the
start? (Hint: use is.na().)
```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, desc(is.na(x)))
```

2. Sort flights to find the most delayed flights. Find the flights
that left earliest.
```{r}
arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)
```
Flight the left the earliest is flight 97 while the most delayed flight is flight 51.

3. Sort flights to find the fastest flights.
```{r}
arrange(flights, air_time)

# another way
arrange(flights, (arr_time - dep_time))
```
Fastest flight is flight 4368 with airtime 20minutes.

4. Which flights traveled the longest? Which traveled the shortest?
```{r}
arrange(flights, distance)
arrange(flights, desc(distance))
```
Flight number 1632 has the shortest distance travelled with 17 miles while flight 51 has the longest distance travelled with 4983 miles.

## Page 54 Exercise

1. Brainstorm as many ways as possible to select dep_time,
dep_delay, arr_time, and arr_delay from flights.
```{r}
select(flights, dep_time,  dep_delay, arr_time, arr_delay)
select(flights, c(dep_time,  dep_delay, arr_time, arr_delay))
flights %>% select(dep_time,  dep_delay, arr_time, arr_delay)
```

2. What happens if you include the name of a variable multiple
times in a select() call?
```{r}
select(flights, dep_time, dep_time, dep_time, dep_time)
```
It will just ignore the multiple variables.

3. What does the one_of() function do? Why might it be helpful
in conjunction with this vector?
vars <- c(
"year", "month", "day", "dep_delay", "arr_delay"
)
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
flights %>% select(one_of(vars))
```
Retuens all variables in the vector for a much cleaner code especially when using `select()`.


4. Does the result of running the following code surprise you?
How do the select helpers deal with case by default? How can
you change that default?
select(flights, contains("TIME"))
```{r}
select(flights, contains("TIME"))
select(flights, contains("TIME", ignore.case = FALSE))
```
The case is insensitive by default w/ the `select()` function. However, we can remedy this using the command`ignore.case = FALSE`.

## Page 58 Exercise

1. Currently dep_time and sched_dep_time are convenient to look
at, but hard to compute with because they’re not really continuous
numbers. Convert them to a more convenient representation
of number of minutes since midnight.
```{r}
# with integer division
mutate(flights, dep_time = (dep_time %/% 100) * 60 + (dep_time %% 100),
       sched_dep_time = (sched_dep_time %/% 100) * 60 + (sched_dep_time %% 100))
```

2. Compare air_time with arr_time - dep_time. What do you
expect to see? What do you see? What do you need to do to fix
it?
```{r}
mutate(flights, dep_time = (dep_time %/% 100) * 60 + (dep_time %% 100),
       sched_dep_time = (sched_dep_time %/% 100) * 60 + (sched_dep_time %% 100),
       arr_time = (arr_time %/% 100) * 60 + (arr_time %% 100),
       sched_arr_time = (sched_arr_time %/% 100) * 60 + (sched_arr_time %% 100)) %>% transmute((arr_time - dep_time) %% (60*24) - air_time)
```
The two variables should be equal but because the new variable take into account the delayed dept, it has higher value. Also, we need to convert `arr_time` and `dep_time` into minutes in order to perform operations to them. `air_time` and `arr_time - dep_time` varies greatly.

3. Compare dep_time, sched_dep_time, and dep_delay. How
would you expect those three numbers to be related?
```{r}
select(flights, dep_time, sched_dep_time, dep_delay)
mutate(flights, 
       dep_time = (dep_time %/% 100)*60 + (dep_time %% 100),
       sched_dep_time = (sched_dep_time %/% 100)*60 + (sched_dep_time %% 100)) %>% transmute(sched_dep_time + dep_delay, dep_time)
```
`sched_dep_time + dep_delay` =`dep_time`, but before that we convert the variables `sched_dep_time` and `dep_delay` into minutes to perform operation.

4. Find the 10 most delayed flights using a ranking function. How
do you want to handle ties? Carefully read the documentation
for min_rank().
```{r}
filter(flights, min_rank(desc(dep_delay))<=10) %>% select(flight, dep_delay)
```

5. What does 1:3 + 1:10 return? Why?
```{r}
1:3 + 1:10 
```
1:10 keeps running, while 1:3 loops.

6. What trigonometric functions does R provide?
```{r}
?Trig
```

## Page 72 Exercise

1. Brainstorm at least five different ways to assess the typical delay
characteristics of a group of flights. Consider the following scenarios:

• A flight is 15 minutes early 50% of the time, and 15 minutes
late 50% of the time.
• A flight is always 10 minutes late.
• A flight is 30 minutes early 50% of the time, and 30 minutes
late 50% of the time.
• 99% of the time a flight is on time. 1% of the time it’s 2 hours
late.

Which is more important: arrival delay or departure delay?
```{r}
str(flights)
head(flights)

flights_delay_summary <- flights %>% group_by(flight) %>%
        summarise(no_of_flights = n(),
                  perentage_of_flights_on_time = sum(arr_time == sched_arr_time)/no_of_flights,
                  percentage_of_late_flights = sum(arr_time > sched_arr_time)/no_of_flights,
                  percentage_of_15_mins_early_flights = sum(sched_arr_time - arr_time == 15)/no_of_flights,
                  percentage_of_15_mins_late_flights = sum(arr_time - sched_arr_time == 15)/no_of_flights, 
                  percentage_of_10_mins_late_flights = sum(arr_time - sched_arr_time == 10)/no_of_flights,
                  percentage_of_30_mins_early_flights = sum(sched_arr_time - arr_time == 30)/no_of_flights,
                  percentage_of_30_mins_late_flights = sum(arr_time - sched_arr_time == 30)/no_of_flights,
                  percentage_of_2hrs_flights = sum(arr_time == 120)/no_of_flights)

flights_delay_summary
```
• A flight is 15 minutes early 50% of the time, and 15 minutes
late 50% of the time.
```{r}
filter(flights_delay_summary, percentage_of_15_mins_early_flights == 0.5 &
       percentage_of_15_mins_late_flights == 0.5)
```
• A flight is always 10 minutes late.
```{r}
filter(flights_delay_summary, percentage_of_10_mins_late_flights > 0) 
```
• A flight is 30 minutes early 50% of the time, and 30 minutes
late 50% of the time.
```{r}
filter(flights_delay_summary, percentage_of_30_mins_early_flights == 0.5 &
       percentage_of_30_mins_late_flights == 0.5)
```

• 99% of the time a flight is on time. 1% of the time it’s 2 hours
late.
```{r}
filter(flights_delay_summary, perentage_of_flights_on_time == .99)
filter(flights_delay_summary, percentage_of_2hrs_flights == .1)
```
2. Come up with another approach that will give you the same
output as not_cancelled %>% count(dest) and not_cancel
led %>% count(tailnum, wt = distance) (without using
count()).
```{r}
# approach used in the example
not_cancelled <- flights %>% filter(!is.na(arr_delay), !is.na(dep_delay))
not_cancelled %>% count(dest)
not_cancelled %>% count(tailnum, wt = distance)

# my approach
not_cancelled %>% group_by(dest) %>%
        summarise(n = n())

not_cancelled %>%
  group_by(tailnum) %>%
  summarise(n = sum(distance))

# dropping all missing values
flights222 <- flights
flights222 %>% drop_na()

# counting all the missing values in the dataframe
sum(is.na(flights))

# missing values of every columns or vars
summary(flights)
```

3. Our definition of cancelled flights (is.na(dep_delay) |
is.na(arr_delay) ) is slightly suboptimal. Why? Which is the
most important column?
```{r}
names(flights)
filter(flights, !is.na(arr_time), !is.na(dep_time))

flights %>%
    group_by(departed = !is.na(dep_delay), arrived = !is.na(arr_delay)) %>%
    summarise(n=n())
```
`arr_time` and `dept_time`.

4. Look at the number of cancelled flights per day. Is there a pattern?
Is the proportion of cancelled flights related to the average
delay?
```{r}
flights %>%
  mutate(dep_date = lubridate::make_datetime(year, month, day)) %>%
        group_by(dep_date) %>% 
        summarise(cancelled = sum(is.na(dep_delay)),
                  n = n(),
                  mean_dep_delay = mean(dep_delay, na.rm = T)) %>% 
        ggplot(aes(x= cancelled/n)) + 
    geom_point(aes(y=mean_dep_delay), colour='blue', alpha=0.5) + 
    ylab('mean delay (minutes)')
```
As we can see, the proportion of cancelled flights is associated with the mean delay although they have no linear relationship.

5. Which carrier has the worst delays? Challenge: can you disentangle
the effects of bad airports versus bad carriers? Why/why
not? (Hint: think about flights %>% group_by(carrier,
dest) %>% summarize(n()).)
```{r}
top10_with_worst_delay <- flights %>% filter(min_rank(desc(dep_delay))<=10) %>% select(carrier)
top10_with_worst_delay
unique(top10_with_worst_delay)

##### not sure w/ this #####
dl <- flights %>% group_by(carrier, dest) %>% summarise(n()) %>% filter(carrier == "DL") %>% top_n(n=5)

flights %>% group_by(carrier, dest) %>% summarize(n()) %>% filter(carrier == "HA") %>% top_n(n=1) 

flights %>% group_by(carrier, dest) %>% summarize(n()) %>% filter(carrier == "MQ") %>% top_n(n=1)

flights %>% group_by(carrier, dest) %>% summarize(n()) %>% filter(carrier == "AA") %>% top_n(n=1)


flights %>% select(carrier, dest, distance) %>% filter(carrier == "HA") %>% top_n(n=1)

flights %>% select(carrier, dest, distance) %>% filter(carrier == "MQ") %>% top_n(n=1)

x <- flights %>% select(carrier, dest, distance) %>% filter(carrier == "AA") 
unique(x) %>% top_n(n=10) %>% arrange(desc(distance))
##### not sure w/ this #####
       
```

HA going to HNL
MQ max RDU
AA DFW
DL ATL

6. For each plane, count the number of flights before the first delay
of greater than 1 hour.
```{r}
# Method 1
flights %>%
    mutate(dep_date = lubridate::make_datetime(year, month, day)) %>%
        group_by(tailnum) %>%
        arrange(dep_date) %>%
        filter(!cumany(arr_delay>60)) %>%
        tally(sort = TRUE)

# Method 2
flights %>%
    mutate(dep_date = lubridate::make_datetime(year, month, day)) %>%
        group_by(tailnum) %>%
        arrange(dep_date) %>%
        filter(cumall(arr_delay<=60)) %>%
        count(sort = TRUE)
```

7. What does the sort argument to count() do? When might you
use it?
The `sort` argument to `count()` sorts by descending order of `n`. This is useful because often the most common group is the most important.

## Page 75 Exercise

1. Refer back to the table of useful mutate and filtering functions.
Describe how each operation changes when you combine it
with grouping.

2. Which plane (tailnum) has the worst on-time record?
```{r}
OnTimeCount <- flights %>% group_by(tailnum) %>%
        filter(arr_delay == 0) %>% tally(sort = T)
OnTimeCount

flights %>% group_by(tailnum) %>% 
        summarise(prop_arrdelay = sum(arr_delay <= 30, na.rm = T)/n(),
                  mean_arrdelay = mean(arr_delay, na.rm = T), flights = n()) %>% arrange(prop_arrdelay, desc(mean_arrdelay))

flights %>%
  group_by(tailnum) %>%
  summarise(prop_on_time = sum(arr_delay <= 30 & !is.na(arr_delay))/n(),
            mean_arr_delay = mean(arr_delay, na.rm=TRUE),
            flights = n()) %>%
  arrange(prop_on_time, desc(mean_arr_delay))

flights %>%
  group_by(tailnum) %>%
  filter(all(is.na(arr_delay))) %>%
    tally(sort=TRUE)
```

3. What time of day should you fly if you want to avoid delays as
much as possible?
```{r}
flights %>% unite(hour, col = time_of_day, hour, minute, sep = "") %>% 
        filter(dep_delay <= 0) %>% group_by(time_of_day) %>% summarise(count = n()) %>% arrange(count)


flights %>%
    ggplot(aes(x=factor(hour), fill=arr_delay>5 | is.na(arr_delay))) + geom_bar()

```
 
4. For each destination, compute the total minutes of delay. For
each flight, compute the proportion of the total delay for its
destination.
```{r}
data1 <- flights %>% group_by(dest) %>% filter(arr_delay > 0) %>% 
        summarise(total_delay = sum(arr_delay))

data2 <- flights %>% group_by(flight, dest) %>% filter(arr_delay > 0) %>%
        summarise(delay = sum(arr_delay, na.rm = T))

data <- merge(data1, data2, by = "dest")
result <- data %>% mutate(prop_delay = delay/total_delay) %>% 
        select(flight, dest, prop_delay, delay, total_delay)
```

5. Delays are typically temporally correlated: even once the problem
that caused the initial delay has been resolved, later flights
are delayed to allow earlier flights to leave. Using lag() explores
how the delay of a flight is related to the delay of the immediately
preceding flight.
```{r}
data <- flights %>% group_by(tailnum, origin) %>%
        filter(dep_delay > 0) %>% 
        arrange(dep_time) %>% 
        summarise(dep_delay, lag = lag(dep_delay), diff = dep_delay - lag)
data <- data[complete.cases(data),]
ggplot(data, mapping = aes(x = lag, y = dep_delay)) + 
        geom_point(color = "Blue")

## Another approach ##
flights %>%
  mutate(new_sched_dep_time = lubridate::make_datetime(year, month, day, hour, minute)) %>%
  group_by(origin) %>%
  arrange(new_sched_dep_time) %>%
  mutate(prev_flight_dep_delay = lag(dep_delay)) %>%
  ggplot(aes(x=prev_flight_dep_delay, y= dep_delay)) + geom_point()
```

We find that there is weak correlation between delays (since no pattern can be observe in the graph).

6. Look at each destination. Can you find flights that are suspiciously
fast? (That is, flights that represent a potential data entry
error.) Compute the air time of a flight relative to the shortest
flight to that destination. Which flights were most delayed in the
air?
```{r}

memo <- flights[complete.cases(flights),]
memo <-  memo %>% group_by(flight, dest) %>% summarise(air_time, fast = min(air_time)) %>% arrange(desc(air_time))
memo
arrange(memo, air_time)

## Another method ##
flights %>%
  mutate(new_sched_dep_time = lubridate::make_datetime(year, month, day, hour, minute)) %>%
  group_by(origin) %>%
  arrange(new_sched_dep_time) %>%
  mutate(prev_flight_dep_delay = lag(dep_delay)) %>%
  lm(dep_delay ~ prev_flight_dep_delay,.) %>% summary()

```
With `air_time` 20 mins, flight 4368 going to BDL is the fastest. On the other hand, the flight 51 and 15 going to `HNL` were most delayed flights with `air_time` 695.

        
7. Find all destinations that are flown by at least two carriers. Use
that information to rank the carriers.
```{r}
flights %>% group_by(dest) %>% 
        filter(n_distinct(carrier) >= 2) %>% 
        group_by(carrier) %>% 
        summarise(rank = n_distinct(dest)) %>% 
        arrange(desc(rank))

```

